---
title: "Introduction to voomCLR"
author: 
  - name: "Koen Van den Berge"
    affiliation:
    - "Janssen Research and Development"
    email: "kvande14@its.jnj.com"
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('voomCLR')`"
vignette: >
  %\VignetteIndexEntry{Introduction to voomCLR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```


```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Track time spent on making the vignette
startTime <- Sys.time()

## Bib setup
library("RefManageR")

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    voomCLR = citation("voomCLR")[1],
    limma = citation("limma")[1]
)
```


# The `voomCLR` workflow

## Example 1: no random effect

Simulate a dataset.

```{r}
set.seed(495212344)
n <- 40 # sample size
P <- 10 # number of cell types
mu0 <- rnbinom(n=P, size=1/2, mu=400)
mu0 # absolute counts in group 0
beta <- rlnorm(n=P, meanlog = 0, sdlog=2) * # these are log-fold-changes
  rbinom(n=P, size=1, prob=.15) *
  sample(c(-1,1), size=P, replace=TRUE) # fold change on log scale
mu1 <- exp(beta) * mu0 # because we want log(mu2/mu1) = beta
relAbundances <- data.frame(g0=mu0/sum(mu0),
                            g1=mu1/sum(mu1)) # relative abundance of absolute count
# relative abundance information (observed data in typical experiment)
Y0 <- rmultinom(n=10, size=1e4, prob=relAbundances$g0)
Y1 <- rmultinom(n=10, size=1e4, prob=relAbundances$g1)
Y <- cbind(Y0, Y1)
rownames(Y) <- paste0("celltype",1:10)
colnames(Y) <- paste0("sample",1:20)
group <- factor(rep(0:1, each=10))
```

The `voomCLR` workflow works much as a usual limma-voom workflow, with two changes:

 - The `voom` function is replaced by a `voomCLR` function.
 - An additional function for performing the bias correction is incorporated: `applyBiasCorrection`.
 
It is essential that the code is run in the particular order as shown below.

```{r}
library(limma)
library(voomCLR)
design <- model.matrix(~group)
v <- voomCLR(counts = Y,
             design = design,
             plot = TRUE,
             span = 0.8) # span for the lowess smoother
fit <- lmFit(v, design)
# plotBeta(fit) # run this to check how confident you can be about the bias correction.
fit <- eBayes(fit)
tt <- topTableBC(fit, coef=2, n=Inf)
head(tt)
```


### Assessing specific contrasts

Sometimes, the contrast of interest may relate to a linear combination of model parameters, rather than a single model parameter. 
In the `limma` framework, we would build contrasts using the `contrasts.fit` function.
Below, we show a simple example.

```{r}
design <- model.matrix(~group)
v <- voomCLR(counts = Y,
             design = design,
             plot = TRUE,
             span = 0.8) # span for the lowess smoother
fit <- lmFit(v, design)
## build the contrast matrix
cm <- matrix(0, nrow=ncol(fit$coefficients), ncol = 1)
rownames(cm) <- colnames(fit$coefficients)
colnames(cm) <- "group1vs0"
cm["group1", "group1vs0"] <- 1
## reparameterize model => one contrast is now one model parameter
contFit <- contrasts.fit(fit, contrasts=cm)
contFit <- eBayes(contFit)
tt <- topTableBC(contFit, coef=1, n=Inf)
head(tt) # note results are identical.
```

## Example 2: approximating a random intercept

you need one extra input: the covariate you'd like to add as random effect.
here I assume this is in the 'patient' object.

```{r}
patient <- rep(paste0("patient",1:10), 2)
v <- voomCLR(counts = Y,
             design = design,
             lib.size = NULL)
## estimate correlation within block
cf <- duplicateCorrelation(v, design, block=patient)
## re-estimate weights conditional on estimated correlation
v <- voomCLR(counts = Y,
             design = design,
             lib.size = NULL,
             block = patient,
             correlation = cf$consensus)
## use 'final' weights to get 'final' estimate of within-block correlation.
cf <- duplicateCorrelation(v, design, block=patient)
fit <- lmFit(v, design,
             block = patient,
             correlation = cf$consensus)
plotBeta(fit)
fit <- eBayes(fit)
tt <- topTableBC(fit, coef=2, n=Inf)
head(tt) 
```






## Citing `voomCLR`

We hope that `r Biocpkg("voomCLR")` will be useful for your research. Please use the following information to cite the package and the overall approach. Thank you!

```{r "citation"}
## Citation info
citation("voomCLR")
```


# Reproducibility

The `r Biocpkg("voomCLR")` package `r Citep(bib[["voomCLR"]])` was made possible thanks to:

* R `r Citep(bib[["R"]])`
* `r Biocpkg("limma")` `r Citep(bib[["limma"]])`
* `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
* `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])`
* `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`
* `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])`
* `r CRANpkg("sessioninfo")` `r Citep(bib[["sessioninfo"]])`
* `r CRANpkg("testthat")` `r Citep(bib[["testthat"]])`

This package was developed using `r BiocStyle::Biocpkg("biocthis")`.


Code for creating the vignette

```{r createVignette, eval=FALSE}
## Create the vignette
library("rmarkdown")
system.time(render("voomCLR.Rmd", "BiocStyle::html_document"))

## Extract the R code
library("knitr")
knit("voomCLR.Rmd", tangle = TRUE)
```

Date the vignette was generated.

```{r reproduce1, echo=FALSE}
## Date the vignette was generated
Sys.time()
```

Wallclock time spent generating the vignette.

```{r reproduce2, echo=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```



# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
with `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])` and `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`.

```{r vignetteBiblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
PrintBibliography(bib, .opts = list(hyperlink = "to.doc", style = "html"))
```
