---
title: "Introduction to voomCLR"
author: 
  - name: "Koen Van den Berge"
    affiliation:
    - "Janssen Research and Development"
    email: "kvande14@its.jnj.com"
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('voomCLR')`"
vignette: >
  %\VignetteIndexEntry{Introduction to voomCLR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  crop = NULL
)
```


```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Track time spent on making the vignette
startTime <- Sys.time()

## Bib setup
library("RefManageR")

## Write bibliography information
bib <- c(
  R = citation(),
  BiocStyle = citation("BiocStyle")[1],
  knitr = citation("knitr")[1],
  RefManageR = citation("RefManageR")[1],
  rmarkdown = citation("rmarkdown")[1],
  sessioninfo = citation("sessioninfo")[1],
  testthat = citation("testthat")[1],
  voomCLR = citation("voomCLR")[1],
  limma = citation("limma")[1]
)
```

# Installation

If installation within `R` using the `install.packages` function does not succeed, you may want to try using the Terminal command `R CMD INSTALL dir`, where `dir` is replaced by the path to the `voomCLR` directory.

# Introduction

The `voomCLR` workflow works much as a usual limma-voom workflow, with two changes:

 - The `voom` function is replaced by a `voomCLR` function. This function transforms the counts using a centered-log-ratio (CLR) transformation, and estimates observation-level heteroscedasticity weights. The latter can be estimated empirically or analytically, as specified using the `varCalc` argument. 
 - The `topTable` function for performing statistical inference is replaced by a `topTableBC` function, where the `BC` stands for bias correction. Before performing inference, the parameter estimates are bias corrected. The uncertainty involved in estimating the bias term can be accommodated using bootstrapping, as specified via the `bootstrap` argument.
 
With respect to the `topTableBC` function, it is important to note that the bias correction only makes sense for testing differences. For example, when testing the difference in expected transformed expression values between groups, as we do below. When testing a baseline mean, e.g. when you are interested in testing whether a feature's average abundance is different from zero (or any other value), the bias correction makes no sense, and one should fall back to the original `topTable` function.
This is because the bias correction leverages that the mode of the parameters should be zero, which comes from assuming that most cell types are not differentially abundant for the specific contrast of interest. Generalizing this to an average abundance test, it makes no sense to assume that most of the cell types in our dataset are not present (i.e., their average abundance is zero).
 

# Transformation, variance weights and model fitting

Simulate a dataset.

```{r}
set.seed(495212344)
n <- 40 # sample size
P <- 10 # number of cell types
mu0 <- rnbinom(n = P, size = 1 / 2, mu = 400)
mu0 # absolute counts in group 0
beta <- rlnorm(n = P, meanlog = 0, sdlog = 2) * # these are log-fold-changes
  rbinom(n = P, size = 1, prob = .15) *
  sample(c(-1, 1), size = P, replace = TRUE) # fold change on log scale
mu1 <- exp(beta) * mu0 # because we want log(mu2/mu1) = beta
relAbundances <- data.frame(
  g0 = mu0 / sum(mu0),
  g1 = mu1 / sum(mu1)
) # relative abundance of absolute count
# relative abundance information (observed data in typical experiment)
Y0 <- rmultinom(n = n/2, size = 1e4, prob = relAbundances$g0)
Y1 <- rmultinom(n = n/2, size = 1e4, prob = relAbundances$g1)
Y <- cbind(Y0, Y1)
rownames(Y) <- paste0("celltype", 1:10)
colnames(Y) <- paste0("sample", 1:20)
group <- factor(rep(0:1, each = 10))
```

## Fixed effects model

```{r}
library(limma)
library(voomCLR)
design <- model.matrix(~group)
v <- voomCLR(
  counts = Y,
  design = design
)
fitFE <- lmFit(v, design)
fitFE <- eBayes(fitFE)
```


## Approximating a random intercept using `limma::duplicateCorrelation`

Here, we approximate a random intercept model using the `patient` variable.

```{r}
patient <- rep(paste0("patient", 1:10), 2)
v <- voomCLR(
  counts = Y,
  design = design,
  lib.size = NULL
)
## estimate correlation within block
cf <- duplicateCorrelation(v, design, block = patient)
## re-estimate weights conditional on estimated correlation
v <- voomCLR(
  counts = Y,
  design = design,
  lib.size = NULL,
  block = patient,
  correlation = cf$consensus
)
## use 'final' weights to get 'final' estimate of within-block correlation.
cf <- duplicateCorrelation(v, design, block = patient)
fit <- lmFit(v, design,
  block = patient,
  correlation = cf$consensus
)
plotBeta(fit) # view distribution of estimated coefficients and their mode
fit <- eBayes(fit)
```

## Analytical heteroscedasticity weights calculation

When there is a limited number of features in your dataset, the empirical mean-variance trend may be highly uncertain, and it could be a better idea to estimate the heteroscedasticity weights analytically.

This can be specified using the `varCalc` argument in the `voomCLR` function. By default, it is set to `"empirical"` which uses the limma-voom implementation of empirically estimating the mean-variance trend. Analytical weights can be calculated by setting `varCalc="analytical"`. By default, we use the negative binomial distribution (`varDistribution="NB"`).

```{r}
v <- voomCLR(
  counts = Y,
  design = design,
  varCalc = "analytical"
)
fitNBW <- lmFit(v, design)
fitNBW <- eBayes(fitNBW)
```


# Statistical inference

The `topTableBC` function works like the original `topTable` function.
Mean model parameters can be specified via a number or character vector.

```{r}
tt1 <- topTableBC(fitFE, coef = 2, n = Inf)
tt2 <- topTableBC(fitFE, coef = "group1", n = Inf)
identical(tt1, tt2)
head(tt1)
```

## Assessing specific contrasts

Sometimes, the contrast of interest may relate to a linear combination of model parameters, rather than a single model parameter. 
In the `limma` framework, we would build contrasts using the `contrasts.fit` function.
Below, we show a simple example.

```{r}
design <- model.matrix(~group)
v <- voomCLR(
  counts = Y,
  design = design
) # span for the lowess smoother
fit <- lmFit(v, design)
## build the contrast matrix
cm <- matrix(0, nrow = ncol(fit$coefficients), ncol = 1)
rownames(cm) <- colnames(fit$coefficients)
colnames(cm) <- "group1vs0"
cm["group1", "group1vs0"] <- 1
## reparameterize model => one contrast is now one model parameter
contFit <- contrasts.fit(fit, contrasts = cm)
contFit <- eBayes(contFit)
tt <- topTableBC(contFit, coef = 1, n = Inf)
head(tt) # note results are identical.
```

## Bootstrapping

The `topTableBC` function implements the bias correction procedure suggested by [Zhou et al. (2022)](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-022-02655-5). Estimation of the bias term involves uncertainty, and we can propagate that uncertainty in the statistical inference via bootstrapping. While both a non-parametric and parametric bootstrap is implemented, we recommend parametric bootstrapping, as it has led to slightly better performances in our evaluations.

Depending on how you have specified the mean model parameter(s) to test, additional arguments may be needed, as we show below.

```{r}
v <- voomCLR(
  counts = Y,
  design = design
) # span for the lowess smoother
fit <- lmFit(v, design)
fit <- eBayes(fit)

## specifying number or character
ttBoot1 <- topTableBC(fit,
  coef = 2, n = Inf,
  bootstrap = "parametric",
  voomWeights = v$weights
)


## using contrast matrix: also provide this matrix,
## specifying the relevant column.
contFit <- contrasts.fit(fit, contrasts = cm)
contFit <- eBayes(contFit)
ttBoot2 <- topTableBC(contFit,
  coef = 1, n = Inf,
  bootstrap = "parametric",
  voomWeights = v$weights,
  contrastMatrix = cm[, 1, drop = FALSE]
)
```


# Citing `voomCLR`

We hope that `r Biocpkg("voomCLR")` will be useful for your research. Please use the following information to cite the package and the overall approach. Thank you!

```{r "citation"}
## Citation info
citation("voomCLR")
```


# Reproducibility

The `r Biocpkg("voomCLR")` package `r Citep(bib[["voomCLR"]])` was made possible thanks to:

* R `r Citep(bib[["R"]])`
* `r Biocpkg("limma")` `r Citep(bib[["limma"]])`
* `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
* `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])`
* `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`
* `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])`
* `r CRANpkg("sessioninfo")` `r Citep(bib[["sessioninfo"]])`
* `r CRANpkg("testthat")` `r Citep(bib[["testthat"]])`

This package was developed using `r BiocStyle::Biocpkg("biocthis")`.


Code for creating the vignette

```{r createVignette, eval=FALSE}
## Create the vignette
library("rmarkdown")
system.time(render("voomCLR.Rmd", "BiocStyle::html_document"))

## Extract the R code
library("knitr")
knit("voomCLR.Rmd", tangle = TRUE)
```

Date the vignette was generated.

```{r reproduce1, echo=FALSE}
## Date the vignette was generated
Sys.time()
```

Wallclock time spent generating the vignette.

```{r reproduce2, echo=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```



# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
with `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])` and `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`.

```{r vignetteBiblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
PrintBibliography(bib, .opts = list(hyperlink = "to.doc", style = "html"))
```
